---
title: "Google Capstone Project [Case Study: How Does a Bike-Share Navigate Speedy Success?]"
author: "Moureen Islam"
date: 01.03.2023
---
## Statement of the Problem



## Data Preparation

**Install the required packages**
```{r}
library(tidyverse)
library(janitor)
library(dplyr)
library(lubridate)
library(ggplot2)
library(geosphere)
```
 **Import raw data**
 
Here are the 12 months' raw data of the year 2022.

```{r}
jan_22 <- read.csv("E:\\Data Analytics\\Capstone Project\\202201-divvy-tripdata.csv") 
                   
feb_22 <- read.csv("E:\\Data Analytics\\Capstone Project\\202202-divvy-tripdata.csv") 
mar_22 <- read.csv("E:\\Data Analytics\\Capstone Project\\202203-divvy-tripdata.csv") 
apr_22 <- read.csv("E:\\Data Analytics\\Capstone Project\\202204-divvy-tripdata.csv") 
may_22 <- read.csv("E:\\Data Analytics\\Capstone Project\\202205-divvy-tripdata.csv") 
june_22 <- read.csv("E:\\Data Analytics\\Capstone Project\\202206-divvy-tripdata.csv") 
july_22 <- read.csv("E:\\Data Analytics\\Capstone Project\\202207-divvy-tripdata.csv") 
aug_22 <- read.csv("E:\\Data Analytics\\Capstone Project\\202208-divvy-tripdata.csv") 
sep_22 <- read.csv("E:\\Data Analytics\\Capstone Project\\202209-divvy-publictripdata.csv") 
oct_22 <- read.csv("E:\\Data Analytics\\Capstone Project\\202210-divvy-tripdata.csv") 
nov_22 <- read.csv("E:\\Data Analytics\\Capstone Project\\202211-divvy-tripdata.csv") 
dec_22 <- read.csv("E:\\Data Analytics\\Capstone Project\\202212-divvy-tripdata.csv")
```

In the following, all the data will be combined into one dataframe. 

```{r}
full_22 <- bind_rows(jan_22, feb_22, mar_22, apr_22, may_22, june_22, july_22, aug_22, sep_22, oct_22, nov_22, dec_22)
colnames(full_22)
```


## Data Processing
 Now, the columns that will not be required for analysis will be removed, and the type of the columns will be checked.

```{r}
full_22 <- full_22 %>%  
  select(-c(ride_id, start_station_name, end_station_name, start_station_id, end_station_id))
str(full_22)

```

**Edit Data**
The day, month, year column will be created for each ride. This will help to aggregate the data for each month or day.

```{r}
full_22$start_date=as.Date(full_22$started_at)
full_22$month=format(as.Date(full_22$start_date, label = TRUE), "%m")
full_22$day=format(as.Date(full_22$start_date), "%d")
full_22$year=format(as.Date(full_22$start_date), "%Y")
full_22$day_name=format(as.Date(full_22$start_date),"%A")
```

To understand the durantion of the ride and the distance travelled, the following calculations will be required.

```{r}
#calculating ride length
full_22$ride_length <- difftime(full_22$ended_at, full_22$started_at, units = "mins")
```

```{r}
#calculate the distance of per ride
full_22 <- full_22 %>%
  mutate(ride_distance = distHaversine(cbind(start_lng, start_lat), cbind(end_lng, end_lat)))
```

```{r}
#check the data again to review the new values

str(full_22)
```
**The values of ride_length is in difftime format. To conduct mathematical calculations, the values need to be converted into numbers.**
```{r}
full_22$ride_length <- as.numeric(full_22$ride_length)
```
**The months are converted into names instead of number.**
```{r}
full_22<- full_22 %>%
  mutate(month = month(started_at, label = TRUE))
```

**Remove the 0s from the new columns**

```{r}
full_22 <- full_22 %>%
  filter(ride_length > 0) %>%
  filter(ride_distance > 0)
```

##Descriptive Analysis

```{r}
##calculate analysis on ride_length and compare between member and casual riders

aggregate(full_22$ride_length ~ full_22$member_casual, FUN = mean)
aggregate(full_22$ride_length ~ full_22$member_casual, FUN = median)
aggregate(full_22$ride_length ~ full_22$member_casual, FUN = max)
aggregate(full_22$ride_length ~ full_22$member_casual, FUN = min)
```
```{r}
#Compare the average ride time by each day for members and casual riders
aggregate(full_22$ride_length ~ full_22$member_casual + full_22$day_name, FUN = mean)

```


## Visual Analysis

**Compare number of rides, member vs casual in the week.**

```{r}
full_22 %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>% 
  group_by(member_casual, weekday) %>% 
  summarise(number_of_rides = n(), average_duration = mean(ride_length)) %>% 
  arrange(member_casual, weekday)  %>% 
  ggplot(aes(x = weekday, y = number_of_rides, fill = member_casual)) +
  geom_col(position = "dodge")

```
```{r}
full_22 %>% 
  group_by(member_casual, month) %>% 
  summarise(number_of_rides = n(), average_duration = mean(ride_length)) %>% 
  arrange(member_casual, month) %>%
  gg
  
  
```



## Save the table to create a dashboard in tableau

```{r}
tripdata_22 <- full_22 %>%
  group_by(member_casual, rideable_type, day_name, month) %>%
  summarise(number_of_rides = n(), avg_ride_length = mean(ride_length), avg_distance = mean(ride_distance))

```

```{r}
write_csv(tripdata_22, "C:\\Users\\MR9\\final-analysis.csv", append = FALSE)

```


