---
title: "Google Capstone Project [Case Study: How Does a Bike-Share Navigate Speedy Success?]"
author: "Moureen Islam"
date: 01.03.2023
---

## Description of the Business Task

This is the final capstone project of "Google Data Analytics Professional Certificate" course. The project is based on a fictional company, Cyclistic and their bike share project. The goal is to attract casual riders to become members. The following data analysis will attempt to answer the following questions-
* How the casual riders and members ride differently?
* Why would casual riders buy Cyclistic annual memberships?

## Data Preparation

The data used are the historic trip data of the year 2022. The source is from:  "https://divvy-tripdata.s3.amazonaws.com/index.html". The data has been made available by Motivate International Inc. This is public data but it has prohibited to access riders' private information.
R and RStudio will be used for data processing and analysis. Tableau will be used for visualization.

**Install the required packages**
```{r}
library(tidyverse)
library(janitor)
library(dplyr)
library(lubridate)
library(ggplot2)
library(geosphere)
```
 **Import raw data**
 
Here are the 12 months' raw data of the year 2022.

```{r}
jan_22 <- read.csv("E:\\Data Analytics\\Capstone Project\\202201-divvy-tripdata.csv") 
                   
feb_22 <- read.csv("E:\\Data Analytics\\Capstone Project\\202202-divvy-tripdata.csv") 
mar_22 <- read.csv("E:\\Data Analytics\\Capstone Project\\202203-divvy-tripdata.csv") 
apr_22 <- read.csv("E:\\Data Analytics\\Capstone Project\\202204-divvy-tripdata.csv") 
may_22 <- read.csv("E:\\Data Analytics\\Capstone Project\\202205-divvy-tripdata.csv") 
june_22 <- read.csv("E:\\Data Analytics\\Capstone Project\\202206-divvy-tripdata.csv") 
july_22 <- read.csv("E:\\Data Analytics\\Capstone Project\\202207-divvy-tripdata.csv") 
aug_22 <- read.csv("E:\\Data Analytics\\Capstone Project\\202208-divvy-tripdata.csv") 
sep_22 <- read.csv("E:\\Data Analytics\\Capstone Project\\202209-divvy-publictripdata.csv") 
oct_22 <- read.csv("E:\\Data Analytics\\Capstone Project\\202210-divvy-tripdata.csv") 
nov_22 <- read.csv("E:\\Data Analytics\\Capstone Project\\202211-divvy-tripdata.csv") 
dec_22 <- read.csv("E:\\Data Analytics\\Capstone Project\\202212-divvy-tripdata.csv")
```

In the following, all the data will be combined into one dataframe. 

```{r}
full_22 <- bind_rows(jan_22, feb_22, mar_22, apr_22, may_22, june_22, july_22, aug_22, sep_22, oct_22, nov_22, dec_22)
colnames(full_22)
```


## Data Processing
 Now, the columns that will not be required for analysis will be removed, and the type of the columns will be checked.

```{r}
full_22 <- full_22 %>%  
  select(-c(ride_id, start_station_name, end_station_name, start_station_id, end_station_id))
str(full_22)

```

**Edit Data**
The day, month, year column will be created for each ride. This will help to aggregate the data for each month or day.

```{r}
full_22$start_date=as.Date(full_22$started_at)
full_22$month=format(as.Date(full_22$start_date, label = TRUE), "%m")
full_22$day=format(as.Date(full_22$start_date), "%d")
full_22$year=format(as.Date(full_22$start_date), "%Y")
full_22$day_name=format(as.Date(full_22$start_date),"%A")
```

To understand the durantion of the ride and the distance travelled, the following calculations will be required.

```{r}
#calculating ride length
full_22$ride_length <- difftime(full_22$ended_at, full_22$started_at, units = "mins")
```

```{r}
#calculate the distance of per ride
full_22 <- full_22 %>%
  mutate(ride_distance = distHaversine(cbind(start_lng, start_lat), cbind(end_lng, end_lat)))
```

```{r}
#check the data again to review the new values

str(full_22)
```
**The values of ride_length is in difftime format. To conduct mathematical calculations, the values need to be converted into numbers.**
```{r}
full_22$ride_length <- as.numeric(full_22$ride_length)
```
**The months are converted into names instead of number.**
```{r}
full_22<- full_22 %>%
  mutate(month = month(started_at, label = TRUE))
```

**Remove the 0s from the new columns**

```{r}
full_22 <- full_22 %>%
  filter(ride_length > 0) %>%
  filter(ride_distance > 0)
```

##Descriptive Analysis

```{r}
##calculate analysis on ride_length and compare between member and casual riders

aggregate(full_22$ride_length ~ full_22$member_casual, FUN = mean)
aggregate(full_22$ride_length ~ full_22$member_casual, FUN = median)
aggregate(full_22$ride_length ~ full_22$member_casual, FUN = max)
aggregate(full_22$ride_length ~ full_22$member_casual, FUN = min)
```
```{r}
#Compare the average ride time by each day for members and casual riders
aggregate(full_22$ride_length ~ full_22$member_casual + full_22$day_name, FUN = mean)

```


## Visual Analysis

**Compare number of rides, member vs casual in the week.**

```{r}
full_22 %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>% 
  group_by(member_casual, weekday) %>% 
  summarise(number_of_rides = n(), average_duration = mean(ride_length)) %>% 
  arrange(member_casual, weekday)  %>% 
  ggplot(aes(x = weekday, y = number_of_rides, fill = member_casual)) +
  geom_col(position = "dodge")

```
```{r}
full_22 %>% 
  group_by(member_casual, month) %>% 
  summarise(number_of_rides = n(), average_duration = mean(ride_length)) %>% 
  arrange(member_casual, month) %>%
  gg
  
  
```



## Save the table to create a dashboard in tableau

```{r}
tripdata_22 <- full_22 %>%
  group_by(member_casual, rideable_type, day_name, month) %>%
  summarise(number_of_rides = n(), avg_ride_length = mean(ride_length), avg_distance = mean(ride_distance))

```

```{r}
write_csv(tripdata_22, "C:\\Users\\MR9\\final-analysis.csv", append = FALSE)

```
## Key Findings

The results show that Casual Riders are more frequent users of bike-share. Weekends are the most used time of the week and May-September are the peak months for bike use.Docked bikes are only used by the casual riders and electric bikes are the most preferable bikes between both type of users. It is also noticed that the average distance for members are almost at the same level for the whole year whereas casual riders used the bikes for long distant travels. So, it can be assumed that the members use cyclistic's bike share for their daily commute, and the casual riders use them for recreation or other sudden commute. 

## Recommendations

* Create marketing campaign to focus on the benefits of membership to engage the casual riders.
* Provide discounts on membership for new members.
* Start special campaigns for summer season as the months are the most popular among the bike riders to use the service.
* Create more flexible payment plans such as: monthly payment or pay per use options for the members.

